\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-08-12 NAR_DRAFT.tex
check in change

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the genome.

>>>>

In order to estimate $\boldsymbol{\pi}$, we generate a subset of variants in the tumor in which we have high confidence that the variants are present, based on an initial calculation of their likelihood give the data.
These are variants such that for any reasonable value of the site-specific prior probability of mutation, the evidence for the variant in the read data erases any doubt about the presence of the variant. 
Let $\mathrm{D}$ be the set of high confidence variant calls, and $\mathrm{s} \in \mathrm{D}$ be the substitution type of each mutation in $\mathrm{D}$.
The posterior distribution of $\boldsymbol{\pi}$ is then $\mathrm{p}(\boldsymbol{\pi} \mid \mathrm{D}) \sim \textrm{Dirichlet}(\boldsymbol{\alpha^{\prime}})$ where
  $$
    \alpha^{\prime}_{m,c} = \alpha_{m,c} + \sum\limits_{\mathrm{s} \in \mathrm{D}} \mathrm{I}\{\mathrm{s} = \mathrm{s}_{m,c}\}.
  $$
Returning to equation \ref{eqn:detailed_prior}, we can compute the posterior predictive probability $p(C = c \mid M,\mathrm{D})$ as
\begin{equation}
  \label{eqn:post_pred}
  $$
  p(C = c \mid M,\mathrm{D}) = \frac{\sum_{m}\alpha^{\prime}_{m,C = c}}{\sum_{m,c}\alpha^{\prime}_{m,c}}.
  $$
\end{equation}
The posterior probability of mutation to allele $m$ given that a mutation has occurred in context $C = c$, $p(m  \mid C = c,\mathrm{D})$ is then
\begin{equation}
  \label{eqn:to_allele}
  $$
   p(m \mid C = c,\mathrm{D}) = \frac{\alpha^{\prime}_{m,c}}{\sum_{m} \alpha^{\prime}_{m,C = c}}.
  $$
\end{equation}
The remaining context dependent component in equation \ref{eqn:detailed_prior} is $p(C = c)$, the prior probability of seeing a particular tri-nucleotide context.
We compute this as simply the proportion of all tri-nucleotide contexts with the context $C = c$ present in the target genome, which is slightly different depending on whether we are analyzing a whole genome or an exome.
